---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# tidypop

<!-- badges: start -->
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![CRAN status](https://www.r-pkg.org/badges/version/tidypop)](https://cran.r-project.org/package=tidypop)
[![Travis build status](https://travis-ci.org/atyre2/tidypop.svg?branch=master)](https://travis-ci.org/atyre2/tidypop)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/atyre2/tidypop?branch=master&svg=true)](https://ci.appveyor.com/project/atyre2/tidypop)
[![Codecov test coverage](https://codecov.io/gh/atyre2/tidypop/branch/dev/graph/badge.svg)](https://codecov.io/gh/atyre2/tidypop?branch=dev)
<!-- badges: end -->


tidy interface to single species population models

The goal of tidypop is twofold. First, to facilitate teaching simple single species population models
in a tidy context. Second, to provide a 'tidy' interface based on data frames to existing population 
packages. 

In a 'tidy' context, the main functions should take inputs and produce outputs that are data frames ready 
to use with `dplyr` and `ggplot2`. Using these packages revolutionized the use of
R in my classroom. Unfortunately, I found myself having to push students much 
farther in the use of programming than I would like in order to use existing 
demographic packages. There was a lot of extra work to convert the return values
into dataframes for plotting purposes. In addition, population models require 
iteration, and that also increased the pressure on their understanding. 

I started [thinking about this a couple of years ago](https://drewtyre.rbind.io/post/teaching-the-logistic-model/)
and this package is the result of that plus many suggestions received on twitter, 
especially from @pixxpih. The key idea is to separate the processing of arguments,
iteration of the model, etc, from the one step transition model. Students can 
easily write a one-line function specifying the transition -- it is often just a
direct translation from a discrete time equation in their notes. 

I also want it to be easy to implement parameters that change with time. Management
changes the parameters of the model, e.g. increasing the per capita birth rate. 
The simplest way to do this is to create a dataframe with one row per time step. 

## Installation

tidypop is not available from [CRAN](https://CRAN.R-project.org) 

but you can install from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("atyre2/tidypop")
```


## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(tidypop)
# a simple exponential growth model
anexppop <- function(N0, b, d){
  N1 <- N0 * (1 + b - d)
  return(N1)
}
# make the input dataframe
inputs <- data.frame(Year = 2018:2025, b = 0.2, d = 0.15)
iterate(inputs, 23, anexppop)
# time dependent model
inputs <- data.frame(Year = 2018:2025, b = seq(0.2, 0.1, length = 8), d = 0.15)
iterate(inputs, 23, anexppop)
```

