---
title: "tidypop examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidypop examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Demonstrate the use of `tidypop` for a range of ecological models, showing how 
to include stochasticity, time dependence, and make plots.

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidypop)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
```

# Exponential growth

`tidypop` uses discrete time, but 
[I think the distinction between exponential and geometric growth is meaningless](https://drewtyre.rbind.io/post/geometric-vs-exponential-growth-models-a-zombie-idea/).
The formula for the model is
$$
N_{t+1} = N_{t}e^{r_t}
$$
and in code:
```{r exponential1a}
exppop <- function(N0, r){
  N1 <- N0*exp(r)
  return(N1)
}
```

`tidypop::iterate()` needs a tibble or dataframe with one row for each timestep.
The name is arbitrary, but I call it parms. There must be one column for each
argument to `exppop()` other than `N0`. It can be useful for plotting to include
some kind of label for each timestep.

```{r exponential1b}
parms <- tibble(t = 1:20, r = rep(0.05, 20))
```

With those two pieces we can now use iterate to add a column with the projected 
population size called N. We also have to choose a starting population size.

```{r exponential1c}
results <- iterate(parms, 33, exppop)
head(results)
```

Making a plot is very straightforward with ggplot

```{r}
ggplot(data = results,
       mapping = aes(x = t, y = N)) +
  geom_point()
```

or with base R

```{r}
plot(N ~ t, data = results)
```

## Stochasticity

To demonstrate how to add stochasticity using `tidypop`, I will use some real data to 
start with.

```{r griz_setup}

grizzly <- data.frame(Year = 1976:2007,
                      N=c(33, 36, 34, 39, 35, 34, 38, 36, 37, 41, 39, 51, 47, 57, 48, 60, 65,
74, 69, 65, 57, 70, 81, 99, 99, 105, 112, 131, 132, 139, 118, 127))
```
These data are a 3-yr running sum of the numbers of
unduplicated grizzly bear females observed with cubs-of-the-year,
1973-2006, in the greater Yellowstone ecosystem (Table 4 from Schwartz,
C.C., M. A. Haroldson, and K. West, eds. 2006. Yellowstone grizzly
bear investigations: annual report of the Interagency Grizzly Bear
Study Team, 2006. US Geological Survey, Bozeman, Montana, USA). First observation is sum of 1973-74-75 counts, second is 1974-75-76, etc.
Compare with Table 3 of Dennis and Taper (1994), which ends at 1991 and
used an unofficial count for 1990 reported by IGBST via phone. Obtained from [Brian Dennis' website on May 13, 2017.](http://www.webpages.uidaho.edu/~brian/reprints/density_dependence_PBLR_test.pdf)


```{r introplot, echo=FALSE}
ggplot(grizzly, aes(x = Year, y = N)) + 
  geom_point() + 
  xlab("Year") + ylab("3 year running sum of females with cubs") 
```

There are 3 sources of variability here: sampling error in population estimation, environmental stochasticity, and demographic stochasticity.

### Sampling error - Epistemic uncertainty

Sampling error means that our estimate of the average rate of growth is uncertain, that is, $\bar{r}$ is a random variable with some variance. One way we can proceed is to assume that *all* of the variation in the graph is sampling error. There is one true value of $r$ at all times, but we don't know what it is. 
So, calculate the per capita rate of population growth for all years, and calculate the mean and standard error. 
$$
r_t = log\left(\frac{n_{t+1}}{n_t}\right)
$$
 
```{r, include=answer}
grizzly <- mutate(grizzly,
       r = log(lead(N)/N))
sumry_griz <- summarize(grizzly,
          rbar = mean(r, na.rm = TRUE),
          r_se = sd(r, na.rm=TRUE)/sqrt(n()-1)) 
# test it out -- assumes you saved summary stats in data_frame called sumry_griz
parms <- tibble(t = 1976:2007,
                r = sumry_griz$rbar)
head(iterate(parms = parms, N0 = 33, popfun = exppop))
```

We need to run the model with many different values of $r$, sampled from the distribution of uncertainty we have for $\bar{r}$.

```{r}
nyears <- length(1976:2007)
bigparms <- tibble(
  t = rep(1976:2007, times = 100),
                  r = rep(rnorm(100, sumry_griz$rbar, sumry_griz$r_se), each = nyears)
                  )
# split() breaks up a dataframe according to the values of f, into a list of smaller dataframe
# map_dfr() is a new function
# map_dfr() is like map but already glues results back into a data frame
bigresults <- split(bigparms, f = bigparms$r) %>% 
  map_dfr(~iterate(parms = ., N0=c(N = 33), popfun = exppop), .id = "rep")

# plot them all
ggplot(data = grizzly,
       mapping = aes(x = Year, y=N)) + 
  geom_line(data = bigresults,
            mapping = aes(x = t, y = N, group=rep),
            alpha = 0.25) + 
  geom_point() 

```

### Environmental stochasticity

We could also assume that all the variation in the graph is caused by variation in birth and death rates between years. Now we also draw random samples of $r$, but we get a new value for each year, and pass that to our model as a vector. 

```{r}
lotsofr <- tibble(rep = rep(1:100, each=nyears),
                  t = rep(1976:2007, times = 100),
                  r = rnorm(100*nyears, sumry_griz$rbar, sumry_griz$r_se))

bigresults2 <- split(lotsofr, f = lotsofr$rep) %>% 
  map(~select(., -rep)) %>% #awkward, sorry
  map_dfr(~iterate(N0=c(N = 33), parms = ., popfun = exppop), .id = "rep")

# plot them all
ggplot(data = grizzly,
       mapping = aes(x = Year, y=N)) + 
  geom_line(data = bigresults2,
            mapping = aes(x = t, y = N, group=rep),
            alpha = 0.25) + 
  geom_point() 
```
 
## Demographic stochasticity

Demographic stochasticity requires more work, because within our function we have to randomly choose how many
individuals live, die, and give birth. We assume births happen before deaths. 

$$
\begin{aligned}
N_{t+1} = & N_{t} + bN_{t}- dN_{t} \\
N_{t+1} = & N_{t}(1 + b - d) \\
N_{t+1} = & N_{t}\lambda = N_{t}e^{r}
\end{aligned}
$$

Here $\lambda$ is the ratio of population size between two times. $log(\lambda) = r$, the intrinsic population growth rate. We will use the first version of the equation, because we want to manipulate $b$ and $d$ separately. 
Births will occur with a Poisson distribution with
mean b*N. Deaths are binomial and occur with a probability $1-e^{-d}$[^1]. 

```{r}
stch_exppop <- function(N0 = 10, t, b = 0.2, d = 0.1){
  births <- rpois(1, b*N0)
  deaths <- rbinom(1, N0, 1-exp(-d))
  N1 <- N0 + births - deaths
  return(N1)
}
```

We can get the mean correct by remembering that $b-d = r$. Before we assumed that $d = 0$, but now we have to 
have a value bigger than zero (more realistic). The average trajectory will be unaffected by the values we pick, 
as long as the difference is `r paste0("$\\bar{r} = ", format(sumry_griz$rbar, digits=3), "$")`
but the variance will change. If we choose $d = 0.2$ then 
`r paste0("$b = ", format(sumry_griz$rbar+0.2, digits=3), "$")`.

```{r}
bigparms2 <- tibble(rep = rep(1:100, each = nyears),
                   t = rep(1976:2007, times = 100),
                   b = 0.2+sumry_griz$rbar,
                   d = 0.2)
# split() breaks up a dataframe according to the values of f, into a list of smaller dataframe
# map_dfr() is a new function
# map_dfr() is like map but already glues results back into a data frame
bigresults2 <- split(bigparms2, f = bigparms2$rep) %>% 
  map(~select(., -rep)) %>% 
  map_dfr(~iterate(N0=c(N = 33), parms = ., popfun = stch_exppop), .id = "rep")

# plot them all
ggplot(data = grizzly,
       mapping = aes(x = Year, y=N)) + 
  geom_line(data = bigresults2,
            mapping = aes(x = t, y = N, group=rep),
            alpha = 0.25) + 
  geom_point() 

```

That looks like the mean is maybe a bit high. Let's check.

```{r}
bigresults2 %>% 
  group_by(rep) %>% 
  mutate(r = log(lead(N)/N)) %>% 
  summarize(rbar = mean(r, na.rm = TRUE),
            r_se = sd(r, na.rm = TRUE)/(n()-1)) %>% 
  ggplot() + geom_histogram(mapping = aes(x = rbar), binwidth = 0.01) + 
  geom_vline(xintercept = sumry_griz$rbar, color = "red", linetype=2)
```

So it is a bit high. That's probably because of "non-linear averaging".

```{r}
nla <- tibble(r = seq(0.0, 0.1, 0.001),
              N32 = 33*exp(r*32))
ggplot(nla,
       mapping = aes(x = r, y = N32)) + geom_line()
```

Population size as a function of *r* is concave up, so if *r* is varying randomly, the average of the growth rates 
is higher than the growth at the average rate. The variation induced by our demographic stochastic model might be too high as well. 

[^1]: Deriving that is a bit complicated. $d$ is the death rate in continuous time. The probability of 0 deaths in
an interval of time is $p(deaths = 0) = e^{-d}$, which is the zero term of a Poisson distribution. We need the probability of at least 1 death so $p(death > 0) = 1 - e^{-d}$ by the law of total probability.
